### 2021-11-24 article.tex
Down to 7 bad boxes...

One notable exception is the TypeScript language, which recently introduced a new feature called \emph{conditional type}, a type-level ternary operator based on subtyping.
A conditional type, written \typescript{S extends T ? Tt : Tf}, reduces to \typescript{Tt} when \typescript{S} is a subtype of \typescript{T}, to \typescript{Tf} when \typescript{S} is not a subtype of \typescript{T}, and is left unreduced when types variables do not allow to draw a conclusion. Unfortunately, the algorithm used to reduce such conditional types is both \emph{unsound} and \emph{incomplete}.
Despite the unsoundness (discussed in \Cref{subsec:conditional-types-in-typescript}), the addition of conditional types to TypeScript illustrates the practical need and timeliness of this feature.

This chapter presents an alternative construct for type-level programming based on subtyping, which we call \emph{match types}.
As the name suggests, match types allow programmers to express types that perform pattern matching on types:
%
\begin{lstlisting}
type Elem[X] = X match {
  case String => Char
  case List[t] => Elem[t]
  case Any => X
}
\end{lstlisting}
%
The example, which we explain in detail in \Cref{sec:overview}, defines the type \dotty{Elem} by matching on the type parameter \dotty{X}.
%
We have implemented match types in the latest version of Scala, an industry-grade, production-ready compiler.
Match types have received a great interest from the Scala community, and are already in active use.

@@

In this section, we offer a brief introduction of match types in Scala by inspecting the example from the previous section in more detail:
%
\begin{lstlisting}
type Elem[X] = X match {
  case String => Char
  case List[t] => Elem[t]
  case Any => X
}
\end{lstlisting}
%
This example defines a type \dotty{Elem} parametrized by one type parameter \dotty{X}. The right-hand side is defined in terms of a match on the type parameter -- a \emph{match type}.
A match type reduces to one of its right-hand sides, depending on the type of its scrutinee.
For example, the above type reduces as follows:
%
\begin{lstlisting}
Elem[String] =:= Char
Elem[Int] =:= Int
Elem[List[Int]] =:= Int
\end{lstlisting}
%
Here we use \dotty{S =:= T} to denote type equality between the two types \dotty{S} and \dotty{T}, witnessed by mutual subtyping.
To reduce a match type, the scrutinee is compared to each pattern, one after the other, using \emph{subtyping}.
For example, although \dotty{String} is a subtype of both \dotty{String} and \dotty{Any} (the top of Scala's subtyping lattice), \dotty{Elem[String]} reduces to \dotty{Char} because the corresponding case appears first.

When the scrutinee type is a \dotty{List}, the match type \dotty{Elem} is defined recursively on the element type of the list.
Hence, in our example \dotty{Elem[List[Int]]} first reduces to the type \dotty{Elem[Int]}, and eventually to the type \dotty{Int}.

\subsection{A Lightweight Form of Dependent Typing}
Match types enable a lightweight form of dependent typing, since term-level pattern matching expressions can be typed accordingly at the type level as a match types.
Consider the following function definition:
%
\begin{lstlisting}
def elem[X](x: X): Elem[X] = x match {
  case x: String => x.charAt(0)
  case x: List[t] => elem(x.head)
  case x: Any => x
}
\end{lstlisting}
%
This definition is well-typed because the match expression in \dotty{elem}'s body has the exact same scrutinee and pattern types as \dotty{Elem[X]} (the function's return type).

Thanks to Scala's type inference, a call to the \dotty{elem} function can have a result type that \emph{depends} on a term-level parameter.
For instance, in the expression \dotty{elem(1)}, the Scala compiler infers the singleton type \dotty{X = 1} for \dotty{elem}'s type parameter.
This expression thus has type \dotty{Elem[1]}, which reduces to \dotty{Int} (via \dotty{Elem}'s third case).
Similarly, in \dotty{elem(x)}, the compiler infers the singleton type \dotty{X = x.type} and the expression has type \dotty{Elem[x.type]}, which might reduce further at the callsite depending on \dotty{x}'s type.

@@

The necessity for disjointness is best illustrated with an example. Consider \dotty{Seq[Int]}, the type of integer sequences. \dotty{Elem[Seq[Int]]} does not reduce:
\begin{enumerate}
  \item \dotty{Elem}'s first case does not apply because \dotty{Seq[Int]} is not a subtype of \dotty{String}.
  \item \dotty{Elem}'s second case does not apply because \dotty{Seq[Int]} is not a subtype of \dotty{List[Int]} (lists are sequences, but not the other way around).
  \item \dotty{Elem}'s third case is \emph{not considered} because \dotty{Seq[Int]} and \dotty{List[Int]} are not disjoint.
\end{enumerate}
Therefore, the reduction algorithm gets stuck on the second case and the overall type is irreducible.
Without disjointness, \dotty{Elem[Seq[Int]]} would reduce to \dotty{Seq[Int]} (via \dotty{Elem}'s third case), which would be unsound. For example, the expression \dotty{elem[Seq[Int]](List(1,2,3))} would have type \dotty{Seq[Int]}, but evaluates to the integer \dotty{1}.

@@

The inheritance and disjointness parameters can be understood as a representation of a hierarchy of Scala traits and classes.
For example, \dotty{trait C1; class C2 extends C1} is represented  in \Fm as $\Ψ \= \{ \(\C_1, \C_2) \};~ \Ξ \= \{ \}$.
This representation also models the fact that certain types cannot possibly have common instances.
For example, \dotty{class C3; class C4} is represented as $\Ψ \= \{ \};~ \Ξ \= \{ \(\C_3, \C_4), \(\C_4, \C_3) \}$, since Scala disallows multiple class inheritance.
Inheritance and disjointness must be consistent in the sense that $\(\A, \B) \∈ \Ξ$ implies that there is no class $\C$ such that $\(\C, \A) \∈ \Ψ$ and $\(\C, \B) \∈ \Ψ$.

@@

% Even though \Fm is limited to type test patterns, more elaborate patterns can be expressed using type tests.
% For example, consider the following data type deconstruction pattern (left-hand side):

% list match {
%   case x :: xs => print(x)
%   case Nil => ()
% }

% list match {
%   case l : ::[Any] => print(l.head)
%   case Nil => ()
% }

% \noindent
% This pattern could alternatively be written as type test pattern (right-hand side), were \dotty{::[Any]} is the type of Scala's cons node.

@@

In this section, we present \SystemFmB, an extension of \SystemFm with support for binding pattern variables in type patterns.
\FmB is parametrized by two sets of classes, $\A$ and $\B$, representing non-parametric and parametric classes, respectively.
A parametric class, written $\B~\T$, takes exactly one type parameter\footnotemark.
We redefine $\C$ to be a syntactic object defined as $\C \::= \A \| \B~\T$.
The class inheritance $\Ψ$ and class disjointness $\Ξ$ remain as binary relations on $\C$.
A generic instantiation in the class hierarchy is represented as an element of $\Ψ$, for example, \dotty{A1 extends B2[A3]} is represented as $\(\A_1, \B_2~\A_3) \∈ \Ψ$.
Generic inheritance is represented using multiple entries in $\Ψ$, for example, \dotty{B1[T] extends B2[T]} is represented as $\∀ \T. \(\B_1~\T, \B_2~\T) \∈ \Ψ$.
This approach allows us to reuse most of \Fm's definitions.
Indeed, our formal development treats $\C$, $\Ψ$, and $\Ξ$ as mathematical objects, and is compatible with \FmB's new definition of classes.

@@

An important limitation of \SystemFm compared to Scala's type system is that it does not support empty types.
The bottom of Scala's subtyping lattice, called \dotty{Nothing}, provides a direct way to refer to empty sets of values.
Intersection types also provide a way to construct uninhabited types given that Scala does not forbid intersecting two disjoint types.
Empty types are problematic for the match type reduction algorithm as they break the fundamental assumption that two types cannot be both disjoint and subtypes (\Cref{lem:disjointness-subtyping-exclusivity}).
To account for this, our implementation uses an additional \emph{inhabitance check} on scrutinee types before attempting any reduction.

To show why empty types are problematic, we can construct an example where, in the absence of an inhabitance check, the same match type could be reduced differently in two different contexts:
%
\begin{lstlisting}
type M[X] = X match {
  case Int => String
  case String => Int
}
class C {
  type X
  def f(bad: M[X & String]): Int = bad
}
class D extends C {
  type X = Int
}
\end{lstlisting}
%
In this example, the definition of \dotty{f} in \dotty{C} type-checks because \dotty{X & String} and \dotty{Int} are disjoint (since \dotty{String} and \dotty{Int} are disjoint) and \dotty{M[X & String]} reduces to \dotty{Int} (\dotty{M}'s second case applies).
Class \dotty{D} refines the definition of \dotty{C} by giving concrete definition of \dotty{X}.
The unsoundness manifests itself in the body of class \dotty{D}, where \dotty{X & String} is a subtype of \dotty{Int} and \dotty{M[X & String]} reduces to \dotty{String} (\dotty{M}'s first case applies).
There, it is possible to call the function~\dotty{f} with a string argument, which would result in a runtime error.
Checking for scrutinee inhabitance prevents this class of errors. In this example, it would prevent \dotty{M[Int & String]} from reducing given that \dotty{Int & String} is not inhabited.

@@

In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \dotty{A | Null}~\citep{nieto2020scala}.
Our implementation of subtyping and disjointness handles union types and therefore needs no particular treatment of null values.

% To allow easy migration from older versions, the strict treatment of nulls is still optional in Scala 3.0, enabled by the command line option \dotty{-Yexplicit-nulls}. The plan is to make strict null checking the default in the future.

@@

Scala supports variance annotations on type parameters of higher-kinded types.
These annotations allow programmers to specify how the subtyping of annotated parameters influences the subtyping of the higher-kinded type.
For instance, \dotty{type F[+T]} defines a type \dotty{F} that is covariant in its first type parameter, meaning that \dotty{T1 <: T2} implies \dotty{F[T1] <: F[T2]}.
Contravariance, written \dotty{type G[-T]}, has the opposite meaning: \dotty{T1 <: T2} implies \dotty{G[T2] <: G[T1]}.

It would appear that co- and contravariant types are always overlapping, given that, for all types \dotty{X}, \dotty{F[Nothing] <:} \dotty{F[X]} and \dotty{G[Any] <: G[X]} (where \dotty{Nothing} and \dotty{Any} are Scala's bottom and top types).
However, in the case of covariant parameters, an exception can be made when the said type parameter corresponds to a field or a constructor parameter: the \dotty{Nothing} instantiation can be ruled out because no runtime program can produce a value of that type.

Scala tuples, for example, fall into this category.
\dotty{Tuple2}, the class for pairs, is defined as follows:
%
\begin{lstlisting}
case class Tuple2[+T1, +T2](_1: T1, _2: T2)
\end{lstlisting}
%
Given two instantiations of this this type, \dotty{Tuple2[Int,} \dotty{X]} and \dotty{Tuple2[String,X]}, although \dotty{Tuple2[Nothing,X]} is a subtype of both, there is no runtime value of type \dotty{Tuple2[} \dotty{Nothing,X]} (since \dotty{Nothing} is uninhabited), and, as a result, those two types are disjoint.
Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.

@@

Erasure directly affects the type tests that can be performed at runtime.
For example, while \dotty{case xs: List[Int] =>} is a syntactically valid pattern, it will lead to a compiler warning since the \dotty{Int} type parameter is eliminated by erasure and cannot be checked at runtime.

@@

\subsection{Shape Errors in Python}
In the example Python code below, the \python{img_batch} ndarray is a batch of 25 randomly generated RGB images of size 256$\×$256.
The code aims to compute a vector of length 25 containing the average grayscale color of each image in the batch, and then create a square 5$\×$5 image of the average grayscale colors.
However, this code contains a shape error and will throw an error at runtime.
%
\begin{lstlisting}[style=py]
import numpy as np
img_batch = np.random.normal(size=(25, 256, 256, 3))
avg_colors = np.mean(img_batch, (0, 1, 2))
avg_color_square = np.reshape(avg_colors, (5, 5))
\end{lstlisting}
%
The error is in the call to \python{np.mean}, which takes as argument a list of axes to reduce along.
Unfortunately, the arguments to \python{np.mean} are off-by-one and result in a vector of length 3 (\python{img_batch}'s last dimension) instead of the intended length of 25 (\python{img_batch}'s first dimension); \python{avg_color} thus contains the average RGB color of the batch instead of the average grayscale color for each image.
The reshape operation will then fail at runtime, as it cannot reshape a 3-element vector into a 25-element matrix.
This error can be difficult to spot, given that NumPy's interface for reducing along multiple axes is index-based.
Runtime errors like this one can be particularly frustrating when they occur late in a long-running computation.

In the remainder of this section, we show how match types can be used to prevent this class of error.
After a preliminary introduction of singleton types (\Cref{subsec:singleton-types}), we introduce ndarray shapes at the type level using HLists (\Cref{subsec:array-shape}).
In \Cref{subsec:shape-match-types}, we show type-level implementations of the \python{np.mean} and \python{np.reshape} operations using match types.
Finally, we show how this newly defined API can detect and report the error from our original Python example.

@@

Scala supports singleton types, which are types inhabited by a single value~\citep{leontiev2014sip}.
For instance, the singleton type \dotty{1} denotes the type containing the integer value \dotty{1}.
The Scala standard library contains several predefined match types to perform arithmetic operations at the type level.
For instance, \dotty{type +[A <: Int,}\dotty{B <: Int]} represents the addition of integer singleton types.
Internally, the compiler is special-cased to implement these arithmetic operations using constant folding.
Representing numbers with singleton types is desirable for practical purposes, but not absolutely necessary for this case study (for instance, Peano numerals could be used instead).

@@

The shape of an ndarray is a list of dimension lengths; we say that the shape $\(\a_{1}, \a_{2}, \dots, \a_n)$ has $\n$ dimensions.
For instance, a three-by-four matrix is a two-dimensional ndarray of shape $\(\3, \4)$.
We represent the shape of an ndarray at the type level using a heterogeneous type list, or HList for short~\citep{kiselyov2004strongly}.
We define an HList called \dotty{Shape} as an ADT with two constructors\footnotemark, \dotty{#:} and \dotty{Ø}.
\footnotetext{%
In the interest of clarity, our presentation omits type bounds that are necessary to guide type inference, for example in the definition of the \dotty{Shape} data type.}
%
\shapeEnum
%
This data type definition allows us to write lists of dimension sizes, both at the term level as \dotty{#:(3, #:(4, Ø))}, and at the type level as \dotty{#:[3, #:[4, Ø]]}.
The HList type can equivalently be written with \dotty{#:} in infix notation, as \dotty{3 #: 4 #: Ø}.

To represent ndarrays, we define the \dotty{NDArray} type. This type is indexed by the ndarray element type (\dotty{T}), and by the ndarray shape (\dotty{S}), represented as an HList:
%
\ndarrayTrait

The goal of our presentation is to define type-safe operations on \dotty{NDArrays}.
Since our focus is on the type-level, we do not include value-level counterparts to \dotty{T} and \dotty{S} in the definition of \dotty{NDArray}, but this would be necessary in a complete implementation.

To construct ndarrays, we define \dotty{random_normal}, which creates an ndarray of a given shape, where all elements are random \dotty{Float}s:\footnotemark
\footnotetext{%
This snippet uses the triple question mark operator, Scala's standard notation for missing or omitted implementations.
}
%
\randomnormalDef

@@

Encoding the types and shapes of ndarrays in the types allows us to readily provide type- and shape-safety for simple ndarray operations.
For instance, the element-wise Hadamard product, written as \python{np.multiply(x, y)}, requires the \python{x} and \python{y} ndarrays to have the same shape and element types.
This constraint does not require any match types, but can simply be expressed as:
%
\multiplyDef
%
However, we will need the additional expressiveness of match types to implement more complex constraints on array shapes, such as for reshapes (\ref{subsubsec:reshape}) and reductions (\ref{subsubsec:reduction}).

@@

An operation commonly used in NumPy is \python{np.reshape}, which changes the shape of an ndarray, but does not change its values.
A restriction imposed by the NumPy API is that the output shape must have the same number of elements as the input shape. The number of elements of a shape is the product of the sizes of its dimensions; an ndarray of shape\hspace{4pt}\dotty{2 #: 3 #: 4 #: Ø}\hspace{4pt}has \begin{math}2\times3\times4 = 24\end{math} elements.
Note that an ndarray of shape \dotty{Ø} is a scalar, and thus has a single element.
This can be naturally expressed with a match type:
%
\numelementsType

To restrict reshaping to be applicable only on valid shapes, the type system must support encoding type equality constraints.
For this, we make use of Scala's implicit parameters, and of the \dotty{=:=} type equality constraint that is a part of Scala's standard library.
%
\reshapeDef
%
With this definition of \dotty{reshape}, the compiler will only accept a usage of \dotty{reshape} if it is able to prove that the number of elements in the old shape is the same as in the new shape.
This example illustrates how match types can be used in concert with existing features like singleton types and implicit resolution to express powerful constraints.

@@

The NumPy API provides a variety of functions to reduce along a set of axes of an ndarray, such as \python{np.mean(ndarray, axes)} or \python{np.var(ndarray, axes)}.
The \python{axes} parameter is a list of indices of dimensions, listing exactly those dimensions that will no longer be present in the output ndarray.
The dimension indices can be unordered and repeated, and out-of-bounds indices result in an error.
In the Python API, passing the \python{None} value instead of a list of axes reduces along all axes, meaning that the operation returns a scalar.
Note that this is the opposite of passing \dotty{Ø}, which means that we reduce over no axes (effectively a no-op).

This behavior is more complex than the previous examples, but can still be described accurately by a match type.
We use a match type called \dotty{Reduce} to compute the return shape of the operation.

@@

We implement reductions along a given list of indices with logic similar to two nested loops.
The outer loop, \dotty{Loop}, traverses the shape and counts the current index.
The inner loops are implemented using \dotty{Contains} and \dotty{Remove}, standard operations on HLists (omitted).
When \dotty{Loop} reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.
%
\reduceType

@@

Having defined \dotty{random_normal}, \dotty{reshape} and \dotty{mean}, we can rewrite our original Python example in Scala:
%
\badreducescala
%
As expected, the call to \dotty{mean} returns a tensor of shape \dotty{3 #: Ø}.
Therefore, the call to \dotty{reshape} does not type-check since the input shape has 3 elements instead of 25.
If we fix the off by one error to reduce along the correct indices, \dotty{1 #: 2 #:} \dotty{3 #: Ø}, the call to \dotty{reshape} type-checks and \dotty{avg_color} has shape \dotty{25 #: Ø}, as expected.

@@

% A significant difference between type-checking type families and match types is that type families are entirely handled during type inference, while match types reduce during subtyping.
% Influencing Haskell's global type inference is notoriously difficult, which can, unfortunately, hinder the usability of type families.
% For instance, in a functional pearl on typed matrices~\citep{santos2020match}, the authors report that their first (simpler) attempt was deemed inconclusive due to limitations of the Haskell type inference when combining type families with pattern matching.

>>>>

One notable exception is the TypeScript language, which recently introduced a new feature called \emph{conditional type}, a type-level ternary operator based on subtyping.
A conditional type, written |S extends T ? Tt : Tf|, reduces to |Tt| when |S| is a subtype of |T|, to |Tf| when |S| is not a subtype of |T|, and is left unreduced when types variables do not allow to draw a conclusion. Unfortunately, the algorithm used to reduce such conditional types is both \emph{unsound} and \emph{incomplete}.
Despite the unsoundness (discussed in \Cref{subsec:conditional-types-in-typescript}), the addition of conditional types to TypeScript illustrates the practical need and timeliness of this feature.

This chapter presents an alternative construct for type-level programming based on subtyping, which we call \emph{match types}.
As the name suggests, match types allow programmers to express types that perform pattern matching on types:
%
\begin{lstlisting}
type Elem[X] = X match {
  case String => Char
  case List[t] => Elem[t]
  case Any => X
}
\end{lstlisting}
%
The example, which we explain in detail in \Cref{sec:overview}, defines the type |Elem| by matching on the type parameter |X|.
%
We have implemented match types in the latest version of Scala, an industry-grade, production-ready compiler.
Match types have received a great interest from the Scala community, and are already in active use.

@@

In this section, we offer a brief introduction of match types in Scala by inspecting the example from the previous section in more detail:
%
\begin{lstlisting}
type Elem[X] = X match {
  case String => Char
  case List[t] => Elem[t]
  case Any => X
}
\end{lstlisting}
%
This example defines a type |Elem| parametrized by one type parameter |X|. The right-hand side is defined in terms of a match on the type parameter -- a \emph{match type}.
A match type reduces to one of its right-hand sides, depending on the type of its scrutinee.
For example, the above type reduces as follows:
%
\begin{lstlisting}
Elem[String] =:= Char
Elem[Int] =:= Int
Elem[List[Int]] =:= Int
\end{lstlisting}
%
Here we use |S =:= T| to denote type equality between the two types |S| and |T|, witnessed by mutual subtyping.
To reduce a match type, the scrutinee is compared to each pattern, one after the other, using \emph{subtyping}.
For example, although |String| is a subtype of both |String| and |Any| (the top of Scala's subtyping lattice), |Elem[String]| reduces to |Char| because the corresponding case appears first.

When the scrutinee type is a |List|, the match type |Elem| is defined recursively on the element type of the list.
Hence, in our example |Elem[List[Int]]| first reduces to the type |Elem[Int]|, and eventually to the type |Int|.

\subsection{A Lightweight Form of Dependent Typing}
Match types enable a lightweight form of dependent typing, since term-level pattern matching expressions can be typed accordingly at the type level as a match types.
Consider the following function definition:
%
\begin{lstlisting}
def elem[X](x: X): Elem[X] = x match {
  case x: String => x.charAt(0)
  case x: List[t] => elem(x.head)
  case x: Any => x
}
\end{lstlisting}
%
This definition is well-typed because the match expression in |elem|'s body has the exact same scrutinee and pattern types as |Elem[X]| (the function's return type).

Thanks to Scala's type inference, a call to the |elem| function can have a result type that \emph{depends} on a term-level parameter.
For instance, in the expression |elem(1)|, the Scala compiler infers the singleton type |X = 1| for |elem|'s type parameter.
This expression thus has type |Elem[1]|, which reduces to |Int| (via |Elem|'s third case).
Similarly, in |elem(x)|, the compiler infers the singleton type |X = x.type| and the expression has type |Elem[x.type]|, which might reduce further at the callsite depending on |x|'s type.

@@

The necessity for disjointness is best illustrated with an example. Consider |Seq[Int]|, the type of integer sequences. |Elem[Seq[Int]]| does not reduce:
\begin{enumerate}
  \item |Elem|'s first case does not apply because |Seq[Int]| is not a subtype of |String|.
  \item |Elem|'s second case does not apply because |Seq[Int]| is not a subtype of |List[Int]| (lists are sequences, but not the other way around).
  \item |Elem|'s third case is \emph{not considered} because |Seq[Int]| and |List[Int]| are not disjoint.
\end{enumerate}
Therefore, the reduction algorithm gets stuck on the second case and the overall type is irreducible.
Without disjointness, |Elem[Seq[Int]]| would reduce to |Seq[Int]| (via |Elem|'s third case), which would be unsound. For example, the expression |elem[Seq[Int]](List(1,2,3))| would have type |Seq[Int]|, but evaluates to the integer |1|.

@@

The inheritance and disjointness parameters can be understood as a representation of a hierarchy of Scala traits and classes.
For example, |trait C1; class C2 extends C1| is represented  in \Fm as $\Ψ \= \{ \(\C_1, \C_2) \};~ \Ξ \= \{ \}$.
This representation also models the fact that certain types cannot possibly have common instances.
For example, |class C3; class C4| is represented as $\Ψ \= \{ \};~ \Ξ \= \{ \(\C_3, \C_4), \(\C_4, \C_3) \}$, since Scala disallows multiple class inheritance.
Inheritance and disjointness must be consistent in the sense that $\(\A, \B) \∈ \Ξ$ implies that there is no class $\C$ such that $\(\C, \A) \∈ \Ψ$ and $\(\C, \B) \∈ \Ψ$.

@@

@@

In this section, we present \SystemFmB, an extension of \SystemFm with support for binding pattern variables in type patterns.
\FmB is parametrized by two sets of classes, $\A$ and $\B$, representing non-parametric and parametric classes, respectively.
A parametric class, written $\B~\T$, takes exactly one type parameter\footnotemark.
We redefine $\C$ to be a syntactic object defined as $\C \::= \A \| \B~\T$.
The class inheritance $\Ψ$ and class disjointness $\Ξ$ remain as binary relations on $\C$.
A generic instantiation in the class hierarchy is represented as an element of $\Ψ$, for example, |A1 extends B2[A3]| is represented as $\(\A_1, \B_2~\A_3) \∈ \Ψ$.
Generic inheritance is represented using multiple entries in $\Ψ$, for example, |B1[T] extends B2[T]| is represented as $\∀ \T. \(\B_1~\T, \B_2~\T) \∈ \Ψ$.
This approach allows us to reuse most of \Fm's definitions.
Indeed, our formal development treats $\C$, $\Ψ$, and $\Ξ$ as mathematical objects, and is compatible with \FmB's new definition of classes.

@@

An important limitation of \SystemFm compared to Scala's type system is that it does not support empty types.
The bottom of Scala's subtyping lattice, called |Nothing|, provides a direct way to refer to empty sets of values.
Intersection types also provide a way to construct uninhabited types given that Scala does not forbid intersecting two disjoint types.
Empty types are problematic for the match type reduction algorithm as they break the fundamental assumption that two types cannot be both disjoint and subtypes (\Cref{lem:disjointness-subtyping-exclusivity}).
To account for this, our implementation uses an additional \emph{inhabitance check} on scrutinee types before attempting any reduction.

To show why empty types are problematic, we can construct an example where, in the absence of an inhabitance check, the same match type could be reduced differently in two different contexts:
%
\begin{lstlisting}
type M[X] = X match {
  case Int => String
  case String => Int
}
class C {
  type X
  def f(bad: M[X & String]): Int = bad
}
class D extends C {
  type X = Int
}
\end{lstlisting}
%
In this example, the definition of |f| in |C| type-checks because |X & String| and |Int| are disjoint (since |String| and |Int| are disjoint) and |M[X & String]| reduces to |Int| (|M|'s second case applies).
Class |D| refines the definition of |C| by giving concrete definition of |X|.
The unsoundness manifests itself in the body of class |D|, where |X & String| is a subtype of |Int| and |M[X & String]| reduces to |String| (|M|'s first case applies).
There, it is possible to call the function~|f| with a string argument, which would result in a runtime error.
Checking for scrutinee inhabitance prevents this class of errors. In this example, it would prevent |M[Int & String]| from reducing given that |Int & String| is not inhabited.

@@

In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \lstinline!A | Null!~\citep{nieto2020scala}.
Our implementation of subtyping and disjointness handles union types and therefore needs no particular treatment of null values.
To allow easy migration from older versions, the strict treatment of nulls is still optional in Scala 3.0, enabled by the command line option |-Yexplicit-nulls|.
The plan is to make strict null checking the default in the future.

@@

Scala supports variance annotations on type parameters of higher-kinded types.
These annotations allow programmers to specify how the subtyping of annotated parameters influences the subtyping of the higher-kinded type.
For instance, |type F[+T]| defines a type |F| that is covariant in its first type parameter, meaning that |T1 <: T2| implies |F[T1] <: F[T2]|.
Contravariance, written |type G[-T]|, has the opposite meaning: |T1 <: T2| implies |G[T2] <: G[T1]|.

It would appear that co- and contravariant types are always overlapping, given that, for all types |X|, |F[Nothing] <: F[X]| and |G[Any] <: G[X]| (where |Nothing| and |Any| are Scala's bottom and top types).
However, in the case of covariant parameters, an exception can be made when the said type parameter corresponds to a field or a constructor parameter: the |Nothing| instantiation can be ruled out because no runtime program can produce a value of that type.

Scala tuples, for example, fall into this category.
|Tuple2|, the class for pairs, is defined as follows:
%
\begin{lstlisting}
case class Tuple2[+T1, +T2](_1: T1, _2: T2)
\end{lstlisting}
%
Given two instantiations of this this type, |Tuple2[Int,X]| and |Tuple2[String,X]|, although |Tuple2[Nothing,X]| is a subtype of both, there is no runtime value of type |Tuple2[Nothing,X]| (since |Nothing| is uninhabited), and, as a result, those two types are disjoint.
Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.

@@

Erasure directly affects the type tests that can be performed at runtime.
For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it will lead to a compiler warning since the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.

@@

\subsection{Shape Errors in Python}
In the example Python code below, the |img_batch| ndarray is a batch of 25 randomly generated RGB images of size 256$\×$256.
The code aims to compute a vector of length 25 containing the average grayscale color of each image in the batch, and then create a square 5$\×$5 image of the average grayscale colors.
However, this code contains a shape error and will throw an error at runtime.
%
\begin{lstlisting}[style=py]
import numpy as np
img_batch = np.random.normal(size=(25, 256, 256, 3))
avg_colors = np.mean(img_batch, (0, 1, 2))
avg_color_square = np.reshape(avg_colors, (5, 5))
\end{lstlisting}
%
The error is in the call to |np.mean|, which takes as argument a list of axes to reduce along.
Unfortunately, the arguments to |np.mean| are off-by-one and result in a vector of length 3 (|img_batch|'s last dimension) instead of the intended length of 25 (|img_batch|'s first dimension); |avg_color| thus contains the average RGB color of the batch instead of the average grayscale color for each image.
The reshape operation will then fail at runtime, as it cannot reshape a 3-element vector into a 25-element matrix.
This error can be difficult to spot, given that NumPy's interface for reducing along multiple axes is index-based.
Runtime errors like this one can be particularly frustrating when they occur late in a long-running computation.

In the remainder of this section, we show how match types can be used to prevent this class of error.
After a preliminary introduction of singleton types (\Cref{subsec:singleton-types}), we introduce ndarray shapes at the type level using HLists (\Cref{subsec:array-shape}).
In \Cref{subsec:shape-match-types}, we show type-level implementations of the |np.mean| and |np.reshape| operations using match types.
Finally, we show how this newly defined API can detect and report the error from our original Python example.

@@

Scala supports singleton types, which are types inhabited by a single value~\citep{leontiev2014sip}.
For instance, the singleton type |1| denotes the type containing the integer value |1|.
The Scala standard library contains several predefined match types to perform arithmetic operations at the type level.
For instance, |type +[A <: Int, B <: Int]| represents the addition of integer singleton types.
Internally, the compiler is special-cased to implement these arithmetic operations using constant folding.
Representing numbers with singleton types is desirable for practical purposes, but not absolutely necessary for this case study (for instance, Peano numerals could be used instead).

@@

The shape of an ndarray is a list of dimension lengths; we say that the shape $\(\a_{1}, \a_{2}, \dots, \a_n)$ has $\n$ dimensions.
For instance, a three-by-four matrix is a two-dimensional ndarray of shape $\(\3, \4)$.
We represent the shape of an ndarray at the type level using a heterogeneous type list, or HList for short~\citep{kiselyov2004strongly}.
We define an HList called |Shape| as an ADT with two constructors\footnotemark, |#:| and |Ø|.
\footnotetext{%
In the interest of clarity, our presentation omits type bounds that are necessary to guide type inference, for example in the definition of the \lstinline!Shape! data type.}
%
\shapeEnum
%
This data type definition allows us to write lists of dimension sizes, both at the term level as |#:(3, #:(4, Ø))|, and at the type level as |#:[3, #:[4, Ø]]|.
The HList type can equivalently be written with |#:| in infix notation, as |3 #: 4 #: Ø|.

To represent ndarrays, we define the |NDArray| type. This type is indexed by the ndarray element type (|T|), and by the ndarray shape (|S|), represented as an HList:
%
\ndarrayTrait

The goal of our presentation is to define type-safe operations on |NDArrays|.
Since our focus is on the type-level, we do not include value-level counterparts to |T| and |S| in the definition of |NDArray|, but this would be necessary in a complete implementation.

To construct ndarrays, we define |random_normal|, which creates an ndarray of a given shape, where all elements are random |Float|s:\footnotemark
\footnotetext{%
This snippet uses the triple question mark operator, Scala's standard notation for missing or omitted implementations.
}
%
\randomnormalDef

@@

Encoding the types and shapes of ndarrays in the types allows us to readily provide type- and shape-safety for simple ndarray operations.
For instance, the element-wise Hadamard product, written as |np.multiply(x, y)|, requires the |x| and |y| ndarrays to have the same shape and element types.
This constraint does not require any match types, but can simply be expressed as:
%
\multiplyDef
%
However, we will need the additional expressiveness of match types to implement more complex constraints on array shapes, such as for reshapes (\ref{subsubsec:reshape}) and reductions (\ref{subsubsec:reduction}).

@@

An operation commonly used in NumPy is |np.reshape|, which changes the shape of an ndarray, but does not change its values.
A restriction imposed by the NumPy API is that the output shape must have the same number of elements as the input shape. The number of elements of a shape is the product of the sizes of its dimensions; an ndarray of shape\hspace{4pt}|2 #: 3 #: 4 #: Ø|\hspace{4pt}has \begin{math}2\times3\times4 = 24\end{math} elements.
Note that an ndarray of shape |Ø| is a scalar, and thus has a single element.
This can be naturally expressed with a match type:
%
\numelementsType

To restrict reshaping to be applicable only on valid shapes, the type system must support encoding type equality constraints.
For this, we make use of Scala's implicit parameters, and of the |=:=| type equality constraint that is a part of Scala's standard library.
%
\reshapeDef
%
With this definition of |reshape|, the compiler will only accept a usage of |reshape| if it is able to prove that the number of elements in the old shape is the same as in the new shape.
This example illustrates how match types can be used in concert with existing features like singleton types and implicit resolution to express powerful constraints.

@@

The NumPy API provides a variety of functions to reduce along a set of axes of an ndarray, such as |np.mean(ndarray, axes)| or |np.var(ndarray, axes)|.
The |axes| parameter is a list of indices of dimensions, listing exactly those dimensions that will no longer be present in the output ndarray.
The dimension indices can be unordered and repeated, and out-of-bounds indices result in an error.
In the Python API, passing the |None| value instead of a list of axes reduces along all axes, meaning that the operation returns a scalar.
Note that this is the opposite of passing |Ø|, which means that we reduce over no axes (effectively a no-op).

This behavior is more complex than the previous examples, but can still be described accurately by a match type.
We use a match type called |Reduce| to compute the return shape of the operation.

@@

We implement reductions along a given list of indices with logic similar to two nested loops.
The outer loop, |Loop|, traverses the shape and counts the current index.
The inner loops are implemented using |Contains| and |Remove|, standard operations on HLists (omitted).
When |Loop| reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.
%
\reduceType
\reduceLoop

@@

Having defined |random_normal|, |reshape| and |mean|, we can rewrite our original Python example in Scala:
%
\badreducescala
%
As expected, the call to |mean| returns a tensor of shape |3 #: Ø|.
Therefore, the call to |reshape| does not type-check since the input shape has 3 elements instead of 25.
If we fix the off by one error to reduce along the correct indices, |1 #: 2 #: 3 #: Ø|, the call to |reshape| type-checks and |avg_color| has shape |25 #: Ø|, as expected.

@@
